Depolarization of interacting NV$^-$ centers at low magnetic fields

### 2022-07-08 Mes papiers/Low field CR/copie.tex
la tete au weekend

\documentclass[preprintnumbers,amsmath,amssymb,superscriptaddress,twocolumn,showpacs]{revtex4-2}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{natbib}
\usepackage{physics}
\usepackage[caption=false]{subfig}

@@

\title{Depolarization of Interacting NV ensembles : \\
The Role of Local Electric Fields and Double Flip processes}

@@

\begin{abstract}
We present experimental results on dipolar interaction processes in ensemble of spin-1 systems. Using the spin of highly doped ensembles of negatively charged nitrogen vacancy centers in diamonds, we identify regimes where flip-flop, double-flip processes as well as the mixing induced by local electric field dominate.
%We analyse these results from the perspective of magnetometry applications, and show that both double-flip processes and local electric fields improve the magnetic field sensitivity.
Our results are relevant for understanding decoherence in many-body spin systems as well as for high sensitivity magneto- and electro-metry with long-lived interacting solid-state spins. As a proof of principle, we present an orientation-free and microwave-free magnetometer operating with a sensitivity below $100$ nT/$\sqrt{\rm Hz}$.
\end{abstract}

@@

The electronic spin properties of the negatively charged nitrogen-vacancy (NV$^-$) center in diamond has given rise to a wealth of applications in nanoscale sensing and quantum information science in part thanks to the possibility to optically polarize and read-out its spin state at ambient conditions. Many recent work indeed focus on ensembles of spins from NV centers with the prospect to boost the magnetic field sensing capabilities and for studying many body effects
\citep{TALLAIRE2020421,edmonds2021characterisation, chatzidrosos2021fiberized, kucsko2018critical, giri_coupled_2018}

When the spin ensemble becomes dense, inhomogeneities in the relaxation rates of NVs can give rise to spin depolarisation with a very rich many body dynamics associated with disorder \citep{choi_observation_2017}. This mechanisms however limits the efficiency of magnetometers \citep{zhou2020quantum}
%It was proposed to exploit these CR to perform sensitive microwave free magnetometry (REF).  
Recent magnetometry proposals have been put forward at zero magnetic field. There, the depolarisation effect was shown to be much stronger, but the detailed studies of relaxation mechanisms have not been identified to the best of our knowledge. Some of the possible processes that lead to cross-relaxation in strongly coupled dipolar systems are depicted in Fig.~1. The goal of this paper is to identify regimes where the spin flip-flop within different classes of NV centers, double-flip up and down as well as the mixing induced by local electric field play a role.

The electronic spin of NV center is a spin-1 system in the ground state (see Fig. 1). It can be optically polarized in the $\ket{m_s=0}$ state. The photoluminescence of this state is also is larger than the $\ket{m_s=\pm 1}$ states enabling spin-read out at ambient temperature. The $\ket{m_s=\pm 1}$ spin states are separated from the $\ket{m_s=0}$ state by  $D = (2\pi) 2.87$ GHz so that when a resonant microwave or static transverse magnetic field is applied, the photoluminescence is reduced \citep{epstein2005anisotropic,lai2009influence}. 
Fig.  \ref{T1}-a) and b) show the change in photoluminescence (PL) with respect to an external magnetic field for two samples with a low and high concentration of NV$^-$ centers (see SI). Both samples show a decrease in PL as the magnetic field amplitude increases.  There is however a stark difference in the low magnetic field region where only high-density samples shows a drop in PL \citep{jarmola_longitudinal_2015,  mrozek_longitudinal_2015}. This effect can be observed on all our samples (see SI) whenever the NV concentration lies in the ppm range. The drop of the PL at low magnetic field is associated with a decrease of the NV's spin lifetime $T_1$ \citep{jarmola_temperature-_2012}, which results in a decrease of the population in the bright state $\ket{0}$.

%Indeed, since the $\ket{m_s=0}$ spin state is brighter than the $\ket{m_s=\pm 1}$ states, and since the optical pumping in the $\ket{m_s=0}$ state is in competition with the relaxation in a thermal distribution of the states, decreasing the spins lifetime results in a decrease of the NV PL \citep{finco2021imaging}.
%To understand the behavior of the PL in zero-field, we therefore need to study the dynamics of the spin in the low magnetic field region.
The depolarization dynamics of the spin of single or dilute NV centers at room temperature is dominated by two-phonon Raman processes \citep{redman1991spin,jarmola_temperature-_2012,norambuena2018spin}, which depend on the crystal lattice temperature. It has been observed that dense NV centers ensemble have an additional spin decay channel \citep{jarmola_temperature-_2012,jarmola_longitudinal_2015,mrozek_longitudinal_2015, choi2017depolarization, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, pellet2021magnetic, mrozek2021characterization}, which depends greatly on the magnetic field amplitude. This effect has been attributed to cross-relaxation between the NV centers through dipole-dipole coupling \citep{mrozek_longitudinal_2015, choi2017depolarization}.  Inhomogeneity of the spin lifetimes is further needed in order to explain the depolarization of the spin ensemble \citep{choi2017depolarization}. We will denote $T_1^{\rm ph}$ the characteristic timescale associated with the phonon relaxation process and $T_1^{\rm dd}$ the density dependent timescale associated with the dipole-dipole relaxation process.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/shema_summary.pdf}
\caption{Schematics showing a diamond with interacting spins as well as the three processes that can account for dipolar relaxation. }
\label{schema_intro}
\end{figure}

@@

\begin{figure*}
\includegraphics[width=.95\textwidth]{Figures/fig_100_vs_1x1x1x1.pdf}
\caption{(a) ODMR spectrum in zero field. (b)-i) ODMR spectrum for a magnetic field $\approx$ 60 G, misaligned by $\sim  24^\circ$ from the [100] axis. (b)-ii) Normalized photoluminescence of the NV$^-$ ensemble as a function of the magnetic field amplitude. (b)-iii) Stretched part of the spin decay $1/T_1^{\rm dd}$ as a function of the magnetic field amplitude. (c)-i), (c)-ii) and (c)-iii) : same measurements but with a magnetic field close to the [100] axis}
\label{100_VS_1x4}
\end{figure*}

Fig. \ref{T1}-c) shows the sequence employed for measuring $T_1^{\rm dd}$ in the dense sample. 
%The spin lifetime protocol used here is described in Fig. \ref{T1} and is based on previous similar experiments 
It consists in a pump-probe measurement where the spins are first polarized in the $\ket{0}$ state by a green laser, and read-out optically after a variable dark time $\tau$. In highly doped samples, this sequence often results in artifacts, mostly due to charge state transfer in the dark \citep{giri_coupled_2018, giri_selective_2019}. It is therefore convenient to repeat the sequence with an additional $\pi$ pulse right before the spin read-out to prepare the remaining $\ket{0}$ polarization into a darker $\ket{+1}$ or $\ket{-1}$ state. By subtracting the result of the two sequences, we select only the spin-dependent part of the signal, with the added benefit of being able to select a specific class of NV centers \citep{jarmola_temperature-_2012, mrozek_longitudinal_2015, choi2017depolarization}.

Fig. \ref{T1}-b) shows the subtracted signals when $B\approx 0$ and $B \approx 50\ \rm G$.
The ensemble spin decay curves are not exponential, which was found to be due to inhomogeneities between the spin lifetimes of dipolar coupled NV$^-$ centers \citep{choi2017depolarization}.
We will base our interpretations of the experimental results on the NV-fluctuator model developed in \citep{choi2017depolarization}. This model postulates the existence of very short-lived NV centers - the so-called fluctuators - which, when coupled resonantly to the other NV centers, act as a sink of polarization for the spin ensemble. A conclusion of this model is that, for a homogeneous 3D distribution of fluctuators, the dipole-induced lifetime should be stretched-exponential with a stretch factor $\beta=1/2$. Such an ensemble lifetime is indeed observed when $T_1^{\rm dd} \ll T_1^{\rm ph}$.
For the samples used in this paper (detailed in SI) $T_1^{\rm dd} \sim T_1^{\rm ph}$, so both decay processes have to be included in the analysis. Fitting the two curves by the product of an exponential and a stretched exponential, we find $T_1^{\rm ph}=3.6$ ms for both and $T_1^{\rm dd}= 0.6\ \rm ms$ and $13.0\ \rm ms$ respectively. We also conducted further measurements that confirm the lifetime-limited character of the fluctuator employed in the model \citep{choi2017depolarization}, see SI. 
%These results show that our sample clearly manifests dipolar interactions amongst closely packed NV centers, but do not tell about the magnitude of the several possible mechanisms involved in the dipolar relaxation. 
%In spin-1 systems, there can indeed be several channels for decay at zero fields such as flip-flop, double flip,

%The resulting fitting formula is therefore :
%\begin{equation}
%S(\tau)=A \exp (-\frac{\tau}{T_1^{\rm ph}} -\sqrt{\frac{\tau}{T_1^{\rm dd}}}),
%\end{equation}
%Further details on the fitting procedure are available in SI. For the rest of the article we will be only interested on the stretched exponential lifetime ${T_1^{\rm dd}}$.

One of the ways to tune dipole-dipole related phenomena is by changing the number of resonant spins. Fig. \ref{100_VS_1x4} (b)-i) shows an optically detected magnetic resonance (ODMR) spectrum from the dense ensemble HPHT-150-1. The observed lines correspond to the projections of the magnetic fields onto the four "classes" of NV centers - that is physical orientation of the NV axis is the diamond crystal cell - together with the two possible spin transitions $\ket{0}\to\ket{-1}$ and $\ket{0}\to\ket{+1}$. The transitions frequencies of the four different classes can be controlled by changing the amplitude and orientation of the external magnetic field. Fig. \ref{100_VS_1x4} (c)-i) shows an example where all four classes are brought to resonance by placing the magnetic field along the [100] crystalline axis. 
%Other example of class resonances are given in the SI.
This configuration will be essential to decipher what dipolar relaxation mechanisms play a role at low field.

Fig. \ref{100_VS_1x4} (b)-ii) and (b)-iii) show the PL from an ensemble of NV centers as a function of the amplitude of a magnetic field aligned as in Fig. \ref{100_VS_1x4} (b)-i). 
We also extract the stretched exponential spin decay time $T_1^{\rm dd}$ by running the sequence shown in Fig. 2-c) at each of the 0.5 G magnetic field values. 
We observe that the 3 \% increase in PL under low magnetic field is indeed correlated with a drop in the spin decay rate from 1.6 kHz to 0.08 kHz, as the magnetic field increases. This result can be explained by the higher number of resonant spins in zero magnetic field, where all four classes are degenerate, as opposed to $\bm B \neq 0$. The half width of the dip ($\approx 10$G) is consistent with a fluctuator model taking into account the change in the spin resonance overlaps as the $B$ field is swept (see SI).
Note that the decrease in PL for $B>40\ \rm G$ is related to state mixing by the transverse magnetic field, so it is not correlated to a modification of the spin lifetime.

Importantly, we observed that the lifted degeneracy of the four NV classes is not the sole reason for the increase of the spin lifetime as the B field increases. Fig. \ref{100_VS_1x4} (c)-ii) and (c)-iii) show the change in the PL and in $1/T_1^{\rm dd}$ as a function of a magnetic field that is aligned along the [100] axis. For this particular orientation, the four classes are always resonant, regardless of the field amplitude. We can see that there still is an increase in the spin decay rate and a corresponding drop in the PL under low magnetic field values, although considerably reduced. 
%Also, the rate of change in both the PL and $T_1^{\rm dd}$ is larger by more than a factor of 3 in the [100] configuration. Although the contrast is smaller, this effect can thus drastically impact zero-field magnetometry protocols based on CR. C'est faux ça, c'est meme plutot le contraire ici
The main aim of this paper is to understand the physics governing the contribution to the spin decay in this second scenario. Note that the slight drop in PL and the corresponding bump for $1/T_1^{\rm dd}$ at $B \sim 20$G is related to dipolar interaction with NV centers that have a $^{13}$C as a first neighbor \cite{pellet2021optical}.

We identified and isolated two additional mechanisms that could explain this effect : local electric fields and double-flip processes. 
The dipole-dipole interaction Hamiltonian between two spins ${\vec S}_1$ and ${\vec S}_2$ reads :
\begin{equation}
\mathcal{H}^{\rm dd}= -\frac{J_0}{r^3}\left(3\left({\vec S}_1 \cdot \vec u \right)\left({\vec S}_2 \cdot \vec u \right) - {\vec S}_1 \cdot {\vec S}_2  \right),
\end{equation}
Where $J_0= (2 \pi) 52\ \rm MHz \cdot \rm{nm}^3$, $\vec r$ is the relative positions of the two spins and $\vec u = \frac{\vec r}{\norm{\vec r}}$. In situations such as the one described in Fig. \ref{100_VS_1x4} (c)(i), the only relevant terms of $\mathcal{H}^{\rm dd}$ in terms of population transfer are the flip-flop terms such as $\mel{0,+1}{\mathcal{H}^{\rm dd}}{+1,0}$. In zero magnetic field however, we have to take into account these two other mechanisms of the dipolar Hamiltonian in consideration.
%where both NV spins involved in the dipole-dipole coupling flip their spin in the same direction.
It was shown in  \cite{mittiga2018imaging} that local electric fields coming from charge traps such as $P_1^+$ and NV$^-$ centers are responsible for the optically detected spin resonance profile at zero magnetic field in large NV ensembles. Magnetic field noise comes in to second order in zero field. %We also do not take into account the hyper-fine structure of the NV center due to the large inhomogeneous broadening of the ESR transitions (see SI). 
We will then consider the following spin Hamiltonian for the electric field dependent NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_{\rm elec}= d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
where $E_{x,y}$ are the projections of local electric fields on the NV axes. 
Under local electric fields with orientations given by the angle $\phi_E={\rm tan}(E_x/E_y)$, the eigenstates of $H_{NV}$ are $\ket{0}$ and $\ket{\pm}=\frac{1}{\sqrt{2}}(\ket{+1}\pm e^{-i\phi_E}\ket{-1})$.
Computing the flip-flop terms $\ket{\pm,0} \bra{0,\pm} $ in the fluctuator model shows that $1/T_1^{dd}$ should be increased at zero field, even after averaging over all possible orientations $\phi_E$ (see SI).  Another mechanism at stake when considering dipolar interactions with spin-1 systems are double-flip processes. 
These processes correspond to terms like $\ket{+1,0} \bra{0,-1}$ in the anisotropic [RQ CLEMENT : pourquoi anisotropic ?] dipolar interaction, giving rise to an exchange of two units of spin-angular momentum. These processes can become significant when the states $\pm 1$ are degenerate and are thus likely to explain part of the zero-field dip in Fig. c)-ii). 
%Theoretical model predicts a factor of $\approx $5 increase in $1/T_1^{dd}$ at zero field.

In order to discriminate the role of these two effects in the low-field region, we performed another similar study under a purely transverse magnetic field.
This regime in fact emulates the effect of the electric field while turning off the resonance condition for double flip processes : indeed for relatively weak transverse magnetic field, the eigen-basis of the spin Hamiltonian is close to $\ket{0},\ket{+},\ket{-}$. This property has been used to observe a linear dependence in the electric field for the spin transition frequencies by applying an external transverse magnetic field \cite{dolde2011electric,qiu2022nanoscale}.
Fig.~\ref{B_transverse} (a) shows the energy levels for the three eigenstates of the NV spin Hamiltonian as a function of a purely transverse magnetic field. We will denote these eigenstates in the general case $\ket{g}$, $\ket{d}$ and $\ket{e}$. For weak transverse magnetic field, $\ket{g} \approx \ket{0}$, $\ket{d}=\ket{-}$ and $\ket{e} \approx \ket{+}$. As the magnetic field increases, $\ket{g}$ and $\ket{e}$ starts to become a mixing of $\ket{0}$ and $\ket{+}$ while $\ket{d}$ remains equal to $\ket{-}$.

>>>>

\documentclass[preprintnumbers,amsmath,amssymb,superscriptaddress,twocolumn,showpacs]{revtex4-2}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage[sort&compress]{natbib}
\usepackage{physics}
\usepackage[caption=false]{subfig}

@@

\title{Depolarization of interacting NV$^-$ centers at low magnetic fields}

@@

\begin{abstract}
We study dipolar interaction processes in dense ensembles of negatively charged nitrogen vacancy (NV$^-$) centers under small magnetic fields. Employing magnetic field scans close to zero-field along specific crystalline direction while recording the NV$^-$ spin relaxation rate, we could identify regimes where flip-flop, double-flip processes as well as mixing induced by local electric field play a role in the NV-NV cross-relaxation .
%We analyse these results from the perspective of magnetometry applications, and show that both double-flip processes and local electric fields improve the magnetic field sensitivity.
Our results are relevant for understanding decoherence in many-body spin systems as well as for high sensitivity magneto- and electro-metry with long-lived interacting solid-state spins. As a proof of principle, we present an orientation and microwave-free magnetometer based on cross-relaxation, and operating with a sensitivity below $100$ nT/$\sqrt{\rm Hz}$.
\end{abstract}

@@

The electronic spin properties of the negatively charged nitrogen-vacancy (NV$^-$) center in diamond has given rise to a wealth of applications in nanoscale sensing and quantum information science thanks in part to the possibility to optically polarize and read-out its spin state at ambient conditions \cite{DOHERTY20131}. In particular, ensembles of NV centers are widely studied for their enhanced magnetic field sensing capabilities \cite{Acosta, TALLAIRE2020421,edmonds2021characterisation, chatzidrosos2021fiberized} and as pristine platforms for studying many body effects \citep{kucsko2018critical, ChoiNat, ZuYao}. When the NV spin concentration reaches ppm values, spin depolarisation, or cross-relaxation (CR) takes place through a very rich many body dynamics associated with disorder \citep{choi_observation_2017}. These mechanisms however limit the efficiency of typical micro-wave based NV magnetometers, but can be harnessed to attain sub-heisenberg limited magnetometers \citep{zhou2020quantum}. Further, cross-relaxation mechanisms can be turned as a tool for magnetic field sensing \cite{akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019}. 
Spectral features in the photoluminescence indeed appear when the magnetic field crosses specific crystal planes where dipolar interactions are enhanced, leading to CR. The projected sensitivity of such magnetometers lie in the tens of pT$/\sqrt{\rm Hz}$  \cite{akhmedzhanov_microwave-free_2017}, on a par with the most sensitive micro-wave based NV magnetometers \cite{Wolf, Sturner}.

%Typical NV magnetometers operate using microwave scans across the spin resonance.  
%It was proposed to exploit these CR to perform sensitive microwave free magnetometry (REF).  
%Recent magnetometry proposals have been put forward at zero magnetic field.

Cross-relaxation features at zero-field have also been observed and could also be deployed for high sensitivity micro-wave free magnetometry.
The CR contrast was indeed shown to be much stronger \citep{jarmola_longitudinal_2015,  mrozek_longitudinal_2015}. However all the relaxation mechanisms at these low field levels have not been identified. 
Here, we study dipolar interaction processes in ensembles of NV centers under small magnetic fields. Specifically, by employing magnetic field scans along specific crystalline directions, we identify regimes where flip-flop, double-flip processes as well as mixing induced by local electric field play a role in the CR. We also present an orientation and microwave-free magnetometer operating with a sensitivity below $100$ nT/$\sqrt{\rm Hz}$ at zero-field, where double-flip processes amongst NV centers dominate.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/shema_summary.pdf}
\caption{Schematics showing a diamond with interacting spins as well as the three processes that can account for dipolar relaxation : flip-flop processes involving two different classes of NV centers, local electric field mixing and double-flip processes where two units of spin angular momentum are exchanged. }
\label{schema_intro}
\end{figure}

The electronic spin of NV center is a spin-1 system in the electronic ground state (see Fig. 1, top left panel), which can be optically polarized in the $\ket{m_s=0}$ state. The photoluminescence of this state is also larger than in the $\ket{m_s=\pm 1}$ states enabling spin-read out at ambient temperature \cite{DOHERTY20131}. The $\ket{m_s=\pm 1}$ spin states are separated from the $\ket{m_s=0}$ state by  $D = (2\pi) 2.87$ GHz so that when a resonant microwave or static transverse magnetic field is applied, the photoluminescence is reduced \citep{epstein2005anisotropic,lai2009influence}.  The processes that lead to cross-relaxation in such a spin-1 system are depicted in Fig.~1. Flip-flop processes involving coupling between NV centers from identical or from different orientations are depicted in the top-right panel. They have already been shown to play a dominant role when tuning $\delta$ {\it via} longitudinal B fields scans, when $B \gtrsim 30$G \cite{choi_observation_2017}. We will show here that under smaller magnetic fields or using purely transverse fields, double-flip up and down as well as the mixing induced by local electric field give a significant contribution to the CR (two bottom panels).

Fig.  \ref{T1}-a) and b) show the change in photoluminescence (PL) as a function of an external magnetic field for two samples with a low and high concentration of NV$^-$ centers respectively (see SI sec. II \citep{SI_low_filed_CR} \nocite{anishchik2015low, filimonenko2020weak, van1990electric}). Both samples show a decrease in PL as the magnetic field amplitude increases.  There is however a stark difference in the low magnetic field region where only high-density samples shows a drop in PL \citep{jarmola_longitudinal_2015,  mrozek_longitudinal_2015}. This effect can be observed on all our samples whenever the NV concentration lies in the ppm range. The drop of the PL at low magnetic field was shown to be associated with a decrease of the NV's spin lifetime $T_1$ \citep{jarmola_temperature-_2012}, which results in a decrease of the population in the bright state $\ket{0}$.

%Indeed, since the $\ket{m_s=0}$ spin state is brighter than the $\ket{m_s=\pm 1}$ states, and since the optical pumping in the $\ket{m_s=0}$ state is in competition with the relaxation in a thermal distribution of the states, decreasing the spins lifetime results in a decrease of the NV PL \citep{finco2021imaging}.
%To understand the behavior of the PL in zero-field, we therefore need to study the dynamics of the spin in the low magnetic field region.
The depolarization dynamics of the NV spins for single or dilute ensembles of NV centers at room temperature is dominated by two-phonon Raman processes \citep{redman1991spin,jarmola_temperature-_2012,norambuena2018spin}, which depend on the crystal lattice temperature. It has been observed that dense NV centers ensemblesin fact  have an additional spin decay channel \citep{jarmola_temperature-_2012,jarmola_longitudinal_2015,mrozek_longitudinal_2015, choi2017depolarization, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, pellet2021magnetic, mrozek2021characterization}, which depends greatly on the magnetic field amplitude. This effect has been attributed to cross-relaxation between the NV centers through dipole-dipole coupling \citep{mrozek_longitudinal_2015, choi2017depolarization}. 
The reason is that at zero-field many NV ``classes" - that is physical orientation of the NV axis is the diamond crystal cell - becomes resonant, which increases the dipolar relaxation.
Inhomogeneity of the spin lifetimes is further needed in order to explain the depolarization of the spin ensemble \citep{choi2017depolarization}. We will denote $T_1^{\rm ph}$ the characteristic timescale associated with the phonon relaxation process and $T_1^{\rm dd}$ the density dependent timescale associated with the dipole-dipole relaxation process.

@@

\begin{figure*}
\includegraphics[width=.95\textwidth]{Figures/fig_100_vs_1x1x1x1.pdf}
\caption{(a) ODMR spectrum in zero field. (b)-i) ODMR spectrum for a magnetic field $\approx$ 60 G, misaligned by $\sim  24^\circ$ from the [100] axis. (b)-ii) Normalized photoluminescence of the NV$^-$ ensemble as a function of the magnetic field amplitude. (b)-iii) Stretched part of the spin decay $1/T_1^{\rm dd}$ as a function of the magnetic field amplitude. (c)-i), (c)-ii) and (c)-iii) : same measurements but with a magnetic field close to the [100] axis. All the measurements were realized using the sample HPHT-150-1. }
\label{100_VS_1x4}
\end{figure*}

Before analyzing in detail the cause of low-field depolarization, we present $T_1$ measurements realized on HPHT-150-1. Fig. \ref{T1}-c) shows the sequence employed for measuring $T_1^{\rm dd}$. 
%The spin lifetime protocol used here is described in Fig. \ref{T1} and is based on previous similar experiments 
It consists in a pump-probe measurement where the spins are first polarized in the $\ket{0}$ state by a green laser, and read-out optically after a variable dark time $\tau$. In highly doped samples, this sequence often results in artifacts, mostly due to charge state transfer in the dark \citep{giri_coupled_2018, giri_selective_2019}. It is therefore convenient to repeat the sequence with an additional $\pi$ pulse right on one of the eight NV spin-resonances before the spin read-out to prepare the remaining $\ket{0}$ polarization into a darker $\ket{+1}$ or $\ket{-1}$ state. By subtracting the result of the two sequences, we select only the spin-dependent part of the signal, with the added benefit of being able to select a specific class of NV centers \citep{jarmola_temperature-_2012, mrozek_longitudinal_2015, choi2017depolarization}.

Fig. \ref{T1}-d) shows the subtracted signals $S_1-S_2$ when $B\approx 0$ and $B \approx 50\ \rm G$.
The relaxation rate is markedly different in both magnetic field settings, with a much more pronounced depolarization at zero-field.
Notably, the ensemble spin decay curves are not exponential. This was found to be due to inhomogeneities between the spin lifetimes of dipolar coupled NV$^-$ centers \citep{choi2017depolarization}.
We will base our interpretations of the experimental results on the NV-fluctuator model developed in \citep{choi2017depolarization}. This model postulates the existence of very short-lived NV centers - the so-called fluctuators - which, when coupled resonantly to the other NV centers, act as a polarization drain for the spin ensemble. A conclusion of this model is that, for a homogeneous 3D distribution of fluctuators, the dipole-induced relaxation should be a stretched-exponential with a stretch factor $\beta=1/2$. Such a trend was indeed observed when $T_1^{\rm dd} \ll T_1^{\rm ph}$ \cite{choi2017depolarization}. We conducted further measurements to confirm the stretch nature of the dipole-induced decay (see SI sec. IV A), as well as confirming the lifetime-limited model of the fluctuators by measuring the resonance condition between classes of NV centers (see SI sec. IV B).
In all dense samples used in this paper (detailed in SI sec. II) $T_1^{\rm dd} \sim T_1^{\rm ph}$, so both decay processes have to be included in the analysis. Fitting the two curves in Fig. 2 d) by the product of a simple and a stretched exponential, we find $T_1^{\rm ph}=3.6$ ms for both curves and $T_1^{\rm dd}= 0.6\ \rm ms$ and $13.0\ \rm ms$ for the $B=0$ and $B=50$G cases respectively. These results thus demonstrates a twenty-fold increase in the dipolar depolarization rate when the B field is turned to zero.
While the fluctuator model developed in \citep{choi2017depolarization} contains some of the explanation for this increase in dipolar depolarization, due to more classes being resonant in zero field than in non-zero field, the model did not include the specificity of the zero magnetic field region, such as the importance of local electric field and the double flip processes. We therefore decided to conducted more in depth experiments in order to characterize these zero field specificities.
%We also conducted further measurements that confirm the lifetime-limited character of the fluctuator employed in the model \citep{choi2017depolarization}, see SI sec. IV B. Using this model and the increase in the single flip-flops rates through the change in the number of resonant spins, we cannot predict such as large change in the relaxation (See SI). We thus conducted more in depth experiments in order to extract the other mechanisms at stake.

%These results show that our sample clearly manifests dipolar interactions amongst closely packed NV centers, but do not tell about the magnitude of the several possible mechanisms involved in the dipolar relaxation. 
%In spin-1 systems, there can indeed be several channels for decay at zero fields such as flip-flop, double flip,

%The resulting fitting formula is therefore :
%\begin{equation}
%S(\tau)=A \exp (-\frac{\tau}{T_1^{\rm ph}} -\sqrt{\frac{\tau}{T_1^{\rm dd}}}),
%\end{equation}
%Further details on the fitting procedure are available in SI. For the rest of the article we will be only interested on the stretched exponential lifetime ${T_1^{\rm dd}}$.
Fig. \ref{100_VS_1x4} (a) shows an optically detected magnetic resonance (ODMR) spectrum at zero magnetic field, from the dense ensemble HPHT-150-1.
Fig. \ref{100_VS_1x4} (b)-i) shows an ODMR spectrum from the same sample, but in a magnetic field. The observed lines correspond to the projections of the magnetic fields onto the four classes of NV centers together with the two possible spin transitions $\ket{0}\to\ket{-1}$ and $\ket{0}\to\ket{+1}$. The transitions frequencies of the four different classes can be controlled by changing the amplitude and orientation of the external magnetic field. Fig. \ref{100_VS_1x4} (c)-i) shows an example where all four classes are brought to resonance by placing the magnetic field along the [100] crystalline axis. 
%Other example of class resonances are given in the SI.
This configuration will be essential to decipher which dipolar relaxation mechanisms play a role at low field.

Fig. \ref{100_VS_1x4} (b)-ii) show the PL from the NV centers ensemble as a function of the amplitude of a magnetic field aligned as in Fig. \ref{100_VS_1x4} (b)-i). 
Fig. \ref{100_VS_1x4} (b)-iii) shows the stretched exponential spin decay time $T_1^{\rm dd}$ obtained using the sequence shown in Fig. 2-c) in the same range of magnetic field values, in 5 G steps. 
We observe that the 3~\% increase in PL under low magnetic field is indeed correlated with a drop in the spin decay rate from 1.6 kHz to 0.08 kHz, as the magnetic field increases. This result can again be explained by the higher number of resonant spins as the zero magnetic field is reduced, where all four classes are degenerate, as opposed to $\bm B \neq 0$. The half width of the dip ($\approx 10$G) is consistent with a fluctuator model taking into account the change in the spin resonance overlaps as the $B$ field is swept (see SI sec. IV B).
Note that the decrease in PL for $B>40\ \rm G$ is related to state mixing by the transverse magnetic field, so it is not associated with a modification of the spin lifetime.

Importantly, B field scans along the [100] direction, where the four classes are always resonant, still features an increase of the spin lifetime as the B field increases. Fig. \ref{100_VS_1x4} (c)-ii) and (c)-iii) show the change in the PL and in $1/T_1^{\rm dd}$ as a function of a magnetic field that is aligned in this direction. We can see that there still is an increase in the spin decay rate and a corresponding drop in the PL under low magnetic field values, although considerably reduced. 
%Also, the rate of change in both the PL and $T_1^{\rm dd}$ is larger by more than a factor of 3 in the [100] configuration. Although the contrast is smaller, this effect can thus drastically impact zero-field magnetometry protocols based on CR. C'est faux ça, c'est meme plutot le contraire ici
The main aim of this paper is to understand the physics governing the contribution to the spin decay in this second scenario. Note that the slight drop in PL and the corresponding bump for $1/T_1^{\rm dd}$ at $B \sim 20$G is related to dipolar interaction with NV centers that have a $^{13}$C as a first neighbor \cite{pellet2021optical}.

We identify and isolate two additional mechanisms that could explain this effect : local electric fields and double-flip processes. 
The dipole-dipole interaction Hamiltonian between two spins ${\vec S}_1$ and ${\vec S}_2$ reads :
\begin{equation}
\mathcal{H}^{\rm dd}= -\frac{J_0}{r^3}\left(3\left({\vec S}_1 \cdot \vec u \right)\left({\vec S}_2 \cdot \vec u \right) - {\vec S}_1 \cdot {\vec S}_2  \right),
\end{equation}
where $J_0= (2 \pi) 52\ \rm MHz \cdot \rm{nm}^3$, $\bm r$ is the relative position of the two spins and $\bm  u = \frac{\bm  r}{\norm{\bm  r}}$. In situations such as the one described in Fig.~\ref{100_VS_1x4} (c)-i), the only relevant terms for depolarisation in $\mathcal{H}^{\rm dd}$ are the flip-flop terms such as $\mel{0,+1}{\mathcal{H}^{\rm dd}}{+1,0}$. In zero magnetic field however, we have to take into account two other mechanisms of the dipolar Hamiltonian in consideration.
%where both NV spins involved in the dipole-dipole coupling flip their spin in the same direction.
It was shown in  \cite{mittiga2018imaging} that local electric fields coming from $P_1^+$ or NV$^-$ centers are responsible for the optically detected spin resonance profile at zero magnetic field in dense NV ensembles. Magnetic field noise comes in to second order in zero field. %We also do not take into account the hyper-fine structure of the NV center due to the large inhomogeneous broadening of the ESR transitions (see SI). 
We will then consider the following spin Hamiltonian for the electric field dependent NV$^-$ ground state: 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_{\rm elec}= d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
where $E_{x,y}$ are the projections of local electric fields on the NV axes. 
Under local electric fields with orientations given by the angle $\phi_E={\rm tan}(E_x/E_y)$, the eigenstates of $H_{NV}$ are $\ket{0}$ and $\ket{\pm}=\frac{1}{\sqrt{2}}(\ket{+1}\pm e^{-i\phi_E}\ket{-1})$.
Computing the flip-flop terms $\ket{\pm,0} \bra{0,\pm} $ in the fluctuator model shows that $1/T_1^{\rm dd}$ should be increased at zero field, even after averaging over all possible orientations $\phi_E$ (see SI sec. V).  Another mechanism at stake when considering dipolar interactions with spin-1 systems are double-flip processes. 
These processes correspond to terms like $\ket{+1,0} \bra{0,-1}$ in the dipolar interaction, giving rise to an exchange of two units of spin-angular momentum. These processes can become significant when the states $\ket{m_s=\pm 1}$ are degenerate and are thus likely to explain part of the zero-field dip in Fig. 3-c)-ii). 
%Theoretical model predicts a factor of $\approx $5 increase in $1/T_1^{dd}$ at zero field.

In order to discriminate the role of these two effects in the low-field region, we make use of a purely transverse magnetic field. The transverse B field emulates the effect of the electric field, while turning off the resonance condition for the effectiveness of double-flip processes. Indeed for relatively weak transverse magnetic field, the eigen-basis of the spin Hamiltonian is close to $\ket{0},\ket{+},\ket{-}$. This property has already been used to observe a linear dependence in the electric field for the spin transition frequencies by applying an external transverse magnetic field \cite{dolde2011electric,qiu2022nanoscale}.
Fig.~\ref{B_transverse} (a) shows the energy levels for the three eigenstates of the NV spin Hamiltonian as a function of a purely transverse magnetic field. We will denote these eigenstates in the general case $\ket{g}$, $\ket{d}$ and $\ket{e}$. For weak transverse magnetic field, $\ket{g} \approx \ket{0}$, $\ket{d}=\ket{-}$ and $\ket{e} \approx \ket{+}$. As the magnetic field increases, $\ket{g}$ and $\ket{e}$ starts to become a mixing of $\ket{0}$ and $\ket{+}$ while $\ket{d}$ remains equal to $\ket{-}$.
